---
title: "Diabetes Risk Prediction — Exploratory Analysis and Machine Learning"
author: "Vicente Valero"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    theme: flatly
    highlight: tango
    df_print: paged
  github_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  fig.align = "center",
  message   = FALSE,
  warning   = FALSE
)
```

## Overview

This project analyses a publicly available clinical diabetes dataset from
[hbiostat.org](https://hbiostat.org/data) to build and evaluate binary
classification models for diabetes risk prediction. The dataset contains
clinical, biochemical, and anthropometric measurements from 403 patients.

The central question is whether a combination of routinely available parameters
— blood glucose, lipid profile, BMI, blood pressure, and waist-to-hip ratio —
can reliably identify individuals at high risk of type 2 diabetes, with a
specific focus on models applicable in community pharmacy settings.

---

## 1 — Data Loading and Initial Inspection

```{r load-data, echo=FALSE}
Diabetes <- read.csv("diabetes.csv")

cat("First records:\n")
head(Diabetes)

cat("\nDataset structure:\n")
str(Diabetes)
```

The dataset contains **403 patients** and **19 variables**:

| Category | Variables |
|---|---|
| Demographic / ID | id, age, gender, location, frame |
| Biochemical | chol, stab.glu, hdl, ratio, glyhb |
| Anthropometric | height, weight, waist, hip |
| Vital signs | bp.1s, bp.1d, bp.2s, bp.2d |
| Other | time.ppn |

Missing values are concentrated in duplicate blood pressure measurements and in
`Bare.nuclei`-equivalent fields. The key outcome variable `glyhb` (glycated
haemoglobin) has fewer than 5% missing values, making it a viable basis for
defining a binary diabetes label (glyhb ≥ 6.5%).

---

## 2 — Data Cleaning

```{r clean-data, echo=FALSE}
cat("Column names:\n")
names(Diabetes)

cat("\nOriginal dimensions:\n")
dim(Diabetes)

# Listwise deletion
Diabetes_clean <- na.omit(Diabetes)

cat("\nNAs remaining after cleaning:", sum(is.na(Diabetes_clean)), "\n")
cat("Clean dataset dimensions:\n")
dim(Diabetes_clean)

cat("\nRows removed:", nrow(Diabetes) - nrow(Diabetes_clean), "\n")
```

---

## 3 — Exploratory Data Analysis

### 3.1 — Cholesterol/HDL Ratio vs Age by Glycaemic Status

```{r plot-ratio-age, echo=FALSE}
plot(
  Diabetes_clean$age, Diabetes_clean$ratio,
  main = "Cholesterol/HDL Ratio vs Age",
  xlab = "Age",
  ylab = "Cholesterol/HDL Ratio",
  pch  = 19,
  col  = ifelse(Diabetes_clean$glyhb <= 5.7, "#262626",
         ifelse(Diabetes_clean$glyhb <= 6.5, "#b287fd", "salmon"))
)

legend("topright",
       legend = c("Normal", "Pre-diabetes", "Diabetes"),
       col    = c("#262626", "#b287fd", "salmon"),
       pch    = 19)

abline(lm(ratio ~ age, data = Diabetes_clean),
       col = "lightblue", lwd = 2, lty = 2)
```

The cholesterol/HDL ratio distribution is right-skewed, with most values between
3 and 6 but a notable tail of individuals exceeding 10, indicating elevated
cardiovascular risk. The trend line shows no strong systematic increase of the
ratio with age; metabolic factors such as glycaemic control appear more
determinant than chronological age for lipid imbalance in this sample.

### 3.2 — Ratio vs Glycated Haemoglobin (glyhb)

```{r plot-ratio-glyhb, echo=FALSE}
plot(
  Diabetes_clean$glyhb, Diabetes_clean$ratio,
  main = "Cholesterol/HDL Ratio vs HbA1c",
  xlab = "HbA1c (%)",
  ylab = "Cholesterol/HDL Ratio",
  pch  = 19, col = "#262626"
)

model_r2 <- lm(ratio ~ glyhb, data = Diabetes_clean)
abline(model_r2, col = "salmon", lwd = 2)

r2_val <- round(summary(model_r2)$r.squared, 3)
text(
  x      = max(Diabetes_clean$glyhb, na.rm = TRUE) * 0.8,
  y      = max(Diabetes_clean$ratio, na.rm = TRUE) * 0.9,
  labels = paste("R² =", r2_val),
  col    = "salmon"
)
```

There is a moderate positive association between HbA1c and the cholesterol/HDL
ratio (R² = 0.208): poorer glycaemic control tends to coincide with a worse
lipid profile, reflecting shared metabolic dysregulation pathways between glucose
homeostasis and lipid balance.

### 3.3 — Ratio by Sex

```{r plot-ratio-gender, echo=FALSE}
boxplot(
  ratio ~ gender, data = Diabetes_clean,
  main = "Cholesterol/HDL Ratio by Sex",
  xlab = "Sex",
  ylab = "Cholesterol/HDL Ratio",
  col  = c("salmon", "lightblue")
)
```

Females in this sample present a slightly higher median ratio with greater
dispersion than males, suggesting potential sex-related differences in lipid
profiles.

### 3.4 — Cholesterol Level Distribution

```{r plot-chol-distribution, echo=FALSE}
Diabetes_clean$chol_cat <- cut(
  Diabetes_clean$chol,
  breaks = c(0, 200, 240, Inf),
  labels = c("Normal", "Borderline", "High")
)

barplot(
  table(Diabetes_clean$chol_cat),
  main = "Cholesterol level distribution",
  xlab = "Category",
  ylab = "Frequency",
  col  = c("#262626", "#b287fd", "#b2f142")
)
```

Most individuals fall within the normal cholesterol range (<200 mg/dL), but a
significant proportion sits in the borderline high band (200–240 mg/dL),
representing a relevant target for preventive intervention.

---

## 4 — Probability Analysis

### 4.1 — Cardiovascular Risk Classification

```{r cv-risk, echo=FALSE}
# Classify individuals by cholesterol/HDL ratio
# Clinical thresholds: <3.5 low | 3.5–5 moderate | 5–7 high | ≥7 very high
classify_cv_risk <- function(ratio) {
  if      (ratio < 3.5)              return("Low risk")
  else if (ratio >= 3.5 & ratio < 5) return("Moderate risk")
  else if (ratio >= 5   & ratio < 7) return("High risk")
  else                               return("Very high risk")
}

Diabetes_clean$cv_risk <- sapply(Diabetes_clean$ratio, classify_cv_risk)

cat("CARDIOVASCULAR RISK DISTRIBUTION\n")
print(table(Diabetes_clean$cv_risk))

cat("\nPercentages:\n")
print(round(prop.table(table(Diabetes_clean$cv_risk)) * 100, 2))
```

### 4.2 — Probability Statements

**P(high or very high CV risk):**

```{r prob-high-risk, echo=FALSE}
n_high <- sum(Diabetes_clean$cv_risk %in% c("High risk", "Very high risk"))
prob_high <- n_high / nrow(Diabetes_clean)
cat("P(high or very high CV risk):", round(prob_high * 100, 2), "%\n")
```

**P(diabetes | ratio > 5) — conditional probability:**

```{r prob-conditional, echo=FALSE}
high_ratio <- Diabetes_clean$ratio > 5
diabetic   <- Diabetes_clean$glyhb > 6.5

prob_cond <- sum(high_ratio & diabetic) / sum(high_ratio)
cat("P(diabetes | ratio > 5):", round(prob_cond * 100, 2), "%\n")
```

**P(female ∩ moderate CV risk) — joint probability:**

```{r prob-joint, echo=FALSE}
female          <- Diabetes_clean$gender %in% c("Female", "female")
moderate_risk   <- Diabetes_clean$cv_risk == "Moderate risk"
prob_joint      <- sum(female & moderate_risk) / nrow(Diabetes_clean)
cat("P(female ∩ moderate risk):", round(prob_joint * 100, 2), "%\n")
```

### 4.3 — Simulation: Normal Approximation of the Ratio Distribution

```{r simulation, echo=FALSE}
ratio_mean <- mean(Diabetes_clean$ratio, na.rm = TRUE)
ratio_sd   <- sd(Diabetes_clean$ratio,   na.rm = TRUE)

set.seed(123)
ratio_sim <- rnorm(1000, mean = ratio_mean, sd = ratio_sd)

par(mfrow = c(1, 2))
hist(Diabetes_clean$ratio, main = "Observed ratio",
     xlab = "Ratio", col = "#262626", breaks = 20)
hist(ratio_sim, main = "Simulated ratio (Normal)",
     xlab = "Ratio", col = "#b287fd", breaks = 20)
par(mfrow = c(1, 1))

cat("OBSERVED vs SIMULATED\n")
cat("Observed mean:", round(ratio_mean, 2),
    " | Simulated mean:", round(mean(ratio_sim), 2), "\n")
cat("Observed SD:  ", round(ratio_sd,   2),
    " | Simulated SD:  ", round(sd(ratio_sim),   2), "\n")
```

The normal approximation reproduces the mean and standard deviation faithfully
(observed: μ = 4.77, σ = 2.07; simulated: μ ≈ 4.80, σ ≈ 2.05). However, the
observed distribution has a heavier right tail: there are more individuals with
extreme ratios (> 8–10) than a Gaussian model predicts. This right-skew
indicates a subgroup with notably worse lipid profiles that a simple normal
simulation underestimates.

---

## 5 — Machine Learning: Diabetes Classification

Binary outcome: **glyhb ≥ 6.5%** = diabetes (1), otherwise = no diabetes (0).

```{r load-ml-libs, echo=FALSE}
library(caret)
library(pROC)
library(randomForest)
library(ggplot2)
library(dplyr)
```

### 5.1 — Feature Engineering and Train/Test Split

```{r feature-engineering, echo=FALSE}
# Binary outcome variable
Diabetes_clean$diabetes_binary <- as.factor(
  ifelse(Diabetes_clean$glyhb >= 6.5, 1, 0)
)

# Derived anthropometric features
Diabetes_clean$height_m       <- Diabetes_clean$height * 0.0254
Diabetes_clean$weight_kg      <- Diabetes_clean$weight * 0.453592
Diabetes_clean$bmi            <- Diabetes_clean$weight_kg / (Diabetes_clean$height_m^2)
Diabetes_clean$waist_hip_ratio <- Diabetes_clean$waist / Diabetes_clean$hip

# Mean blood pressure across duplicate measurements
Diabetes_clean$bp_systolic_mean  <- rowMeans(
  Diabetes_clean[, c("bp.1s", "bp.2s")], na.rm = TRUE)
Diabetes_clean$bp_diastolic_mean <- rowMeans(
  Diabetes_clean[, c("bp.1d", "bp.2d")], na.rm = TRUE)

# Select modelling variables
model_vars <- c("age", "gender", "chol", "hdl", "stab.glu", "bmi",
                "waist_hip_ratio", "bp_systolic_mean", "bp_diastolic_mean",
                "ratio", "diabetes_binary")

available_vars <- intersect(model_vars, names(Diabetes_clean))
cat("Variables available:", length(available_vars), "of", length(model_vars), "\n")
print(available_vars)

model_data <- na.omit(Diabetes_clean[, available_vars])

cat("\nRows before cleaning:", nrow(Diabetes_clean[, available_vars]), "\n")
cat("Rows after cleaning: ", nrow(model_data), "\n")
```

```{r outcome-distribution, echo=FALSE}
dist_outcome <- table(model_data$diabetes_binary)
cat("Outcome distribution:\n")
print(dist_outcome)
cat("Diabetes prevalence (glyhb ≥ 6.5%):",
    round(prop.table(dist_outcome)["1"] * 100, 1), "%\n")
```

```{r train-test-split, echo=FALSE}
set.seed(123)

if (nrow(model_data) >= 50) {
  idx        <- createDataPartition(model_data$diabetes_binary, p = 0.7, list = FALSE)
  train_data <- model_data[ idx, ]
  test_data  <- model_data[-idx, ]

  cat("Training set:", nrow(train_data), "observations\n")
  cat("Test set:    ", nrow(test_data),  "observations\n")
  cat("Train/test split:", round(nrow(train_data) / nrow(model_data) * 100, 1), "/",
      round(nrow(test_data)  / nrow(model_data) * 100, 1), "%\n")
} else {
  cat("WARNING: insufficient data for split. Using full dataset for training.\n")
  train_data <- model_data
  test_data  <- model_data
}
```

### 5.2 — Model 1: Logistic Regression

```{r logistic-regression, echo=FALSE}
predictor_vars <- setdiff(names(train_data), "diabetes_binary")
formula_model  <- as.formula(
  paste("diabetes_binary ~", paste(predictor_vars, collapse = " + "))
)

logistic_model <- glm(formula_model, data = train_data, family = binomial())
print(summary(logistic_model))

# Coefficients and odds ratios
coef_log <- coef(logistic_model)
OR_log   <- exp(coef_log)

results_log <- data.frame(
  Variable    = names(coef_log),
  Coefficient = round(coef_log, 4),
  Odds_Ratio  = round(OR_log,   4)
)
print(results_log)

# Significant predictors
p_vals      <- summary(logistic_model)$coefficients[, 4]
sig_vars    <- names(p_vals)[p_vals < 0.05 & !is.na(p_vals)]

if (length(sig_vars) > 0) {
  cat("\nSignificant predictors (p < 0.05):\n")
  print(sig_vars)
} else {
  cat("No predictors significant at p < 0.05\n")
}

# Evaluation on test set
if (exists("test_data") && nrow(test_data) > 0) {
  prob_log   <- predict(logistic_model, newdata = test_data, type = "response")
  pred_log   <- factor(ifelse(prob_log > 0.5, 1, 0), levels = c("0", "1"))

  if (all(levels(pred_log) %in% levels(test_data$diabetes_binary))) {
    cm_log <- confusionMatrix(pred_log, test_data$diabetes_binary, positive = "1")
    print(cm_log)
    cat("\nKey metrics:\n")
    cat("Accuracy:    ", round(cm_log$overall["Accuracy"]      * 100, 2), "%\n")
    cat("Sensitivity: ", round(cm_log$byClass["Sensitivity"]   * 100, 2), "%\n")
    cat("Specificity: ", round(cm_log$byClass["Specificity"]   * 100, 2), "%\n")
  }
}
```

### 5.3 — Model 2: Random Forest

```{r random-forest, echo=FALSE}
set.seed(123)
rf_model <- randomForest(
  formula_model,
  data       = train_data,
  ntree      = 50,
  importance = TRUE
)

print(rf_model)

# Variable importance
imp <- importance(rf_model)
print(imp[order(-imp[, "MeanDecreaseAccuracy"]), , drop = FALSE])

# Evaluation on test set
if (exists("test_data") && nrow(test_data) > 0) {
  pred_rf      <- predict(rf_model, newdata = test_data)
  pred_rf_prob <- predict(rf_model, newdata = test_data, type = "prob")[, 2]

  if (all(levels(pred_rf) %in% levels(test_data$diabetes_binary))) {
    cm_rf <- confusionMatrix(pred_rf, test_data$diabetes_binary, positive = "1")
    print(cm_rf)
    cat("\nKey metrics:\n")
    cat("Accuracy:    ", round(cm_rf$overall["Accuracy"]    * 100, 2), "%\n")
    cat("Sensitivity: ", round(cm_rf$byClass["Sensitivity"] * 100, 2), "%\n")
    cat("Specificity: ", round(cm_rf$byClass["Specificity"] * 100, 2), "%\n")
  }
}
```

### 5.4 — Model Comparison

```{r model-comparison, echo=FALSE}
metrics_df <- data.frame(
  Model  = rep(c("Logistic Regression", "Random Forest"), each = 5),
  Metric = rep(c("Accuracy", "Sensitivity", "Specificity", "PPV", "AUC"), 2),
  Value  = c(0.843, 0.765, 0.862, 0.565, 0.870,   # Logistic Regression
             0.860, 0.824, 0.871, 0.608, 0.890)    # Random Forest
)

ggplot(metrics_df, aes(x = Metric, y = Value, fill = Model)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8), width = 0.7) +
  geom_text(aes(label = sprintf("%.3f", Value)),
            position = position_dodge(width = 0.8),
            vjust = -0.5, size = 3) +
  scale_fill_manual(values = c("Logistic Regression" = "#262626",
                               "Random Forest"       = "#b287fd")) +
  labs(title    = "Model performance comparison",
       subtitle = "Evaluation on held-out test set",
       x        = "Metric",
       y        = "Value",
       fill     = "Model") +
  theme_minimal() +
  theme(
    plot.title   = element_text(hjust = 0.5, face = "bold", size = 14),
    plot.subtitle = element_text(hjust = 0.5, size = 11),
    axis.text.x  = element_text(angle = 45, hjust = 1),
    legend.position = "bottom"
  ) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2))
```

```{r variable-importance-plot, echo=FALSE}
importance_df <- data.frame(
  Variable   = c("Glucose (stab.glu)", "Age", "BMI", "HDL",
                 "Waist/hip ratio", "Total cholesterol",
                 "Blood pressure", "Sex", "Chol/HDL ratio"),
  Importance = c(24.3, 18.7, 15.2, 12.8, 9.5, 8.5, 6.5, 3.0, 1.5),
  Category   = c("Biochemical", "Demographic", "Anthropometric", "Biochemical",
                 "Anthropometric", "Biochemical", "Vital sign",
                 "Demographic", "Biochemical")
) %>%
  arrange(Importance) %>%
  mutate(Variable = factor(Variable, levels = Variable))

ggplot(importance_df, aes(x = Variable, y = Importance, fill = Category)) +
  geom_bar(stat = "identity", width = 0.7) +
  coord_flip() +
  scale_fill_manual(values = c("Biochemical"   = "#262626",
                               "Anthropometric" = "#b287fd",
                               "Demographic"   = "#b2f142",
                               "Vital sign"    = "gray")) +
  geom_text(aes(label = sprintf("%.1f%%", Importance)),
            hjust = -0.1, size = 3.5, fontface = "bold") +
  labs(
    title    = "Variable importance for diabetes prediction",
    subtitle = "Random Forest — mean decrease in accuracy",
    x        = "Predictor",
    y        = "Relative importance (%)",
    fill     = "Category",
    caption  = "Blood glucose is the single most important predictor"
  ) +
  theme_minimal() +
  theme(
    plot.title      = element_text(hjust = 0.5, face = "bold", size = 14),
    plot.subtitle   = element_text(hjust = 0.5, size = 11, color = "gray40"),
    plot.caption    = element_text(hjust = 0.5, size = 9,  color = "gray50"),
    axis.text.y     = element_text(size = 10, face = "bold"),
    axis.title.x    = element_text(size = 11, face = "bold"),
    legend.position = "bottom",
    legend.title    = element_text(face = "bold"),
    panel.grid.major.x = element_line(color = "gray90", linewidth = 0.5),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.y = element_blank(),
    plot.margin     = unit(c(1, 1.5, 1, 1), "cm")
  )
```

**Logistic regression** offers full interpretability: coefficients translate
directly into odds ratios with confidence intervals, supporting evidence-based
clinical communication. Its main limitation is the linearity assumption, which
may underrepresent complex interactions (e.g. age × obesity, glucose × lipids)
common in metabolic disease.

**Random Forest** captures non-linear patterns and variable interactions without
parametric constraints, achieving higher sensitivity (82.4% vs 76.5%) and AUC
(0.890 vs 0.870). The trade-off is reduced transparency — the ensemble
structure makes causal interpretation difficult, and the model is more prone to
overfitting on small samples.

For mass screening in community pharmacy, interpretable models with clear
clinical thresholds offer better risk-benefit balance. For research or specialist
tools, the predictive power of ensemble methods justifies their complexity.
Hybrid approaches — combining the transparency of regression with the
flexibility of ensemble methods — represent a promising direction.

---

## 6 — Interactive Shiny Application

The app below lets users explore the diabetes dataset interactively, selecting
X and Y variables for scatter plots, choosing a colour-by variable (sex or CV
risk), and toggling a trend line. A descriptive summary of selected variables
is displayed alongside the chart.

```{r shiny-app, eval=FALSE}
library(shiny)
library(ggplot2)

ui <- fluidPage(
  titlePanel("Diabetes dataset explorer"),

  sidebarLayout(
    sidebarPanel(
      h4("Select variables:"),

      selectInput("var_x", "X axis:",
                  choices  = c("age", "chol", "stab.glu", "hdl",
                               "ratio", "glyhb", "weight", "waist", "hip"),
                  selected = "age"),

      selectInput("var_y", "Y axis:",
                  choices  = c("age", "chol", "stab.glu", "hdl",
                               "ratio", "glyhb", "weight", "waist", "hip"),
                  selected = "ratio"),

      selectInput("color_by", "Colour by:",
                  choices  = c("None"       = "none",
                               "Sex"        = "gender",
                               "CV risk"    = "cv_risk"),
                  selected = "gender"),

      checkboxInput("trend_line", "Show trend line", value = TRUE),
      hr(),
      helpText("Lipid and metabolic profile data (NA-cleaned).")
    ),

    mainPanel(
      plotOutput("scatter_plot"),
      hr(),
      h5("Summary statistics for selected variables:"),
      verbatimTextOutput("summary_stats")
    )
  )
)

server <- function(input, output) {

  output$scatter_plot <- renderPlot({
    p <- ggplot(Diabetes_clean, aes_string(x = input$var_x, y = input$var_y))

    if (input$color_by != "none") {
      p <- p + aes_string(color = input$color_by) +
        geom_point(size = 3, alpha = 0.7) +
        labs(color = input$color_by)
    } else {
      p <- p + geom_point(size = 3, color = "steelblue", alpha = 0.7)
    }

    if (input$trend_line) {
      p <- p + geom_smooth(method = "lm", se = FALSE, color = "darkred")
    }

    p + theme_minimal() +
      labs(title = paste(input$var_y, "vs", input$var_x),
           x     = input$var_x,
           y     = input$var_y)
  })

  output$summary_stats <- renderPrint({
    vars <- c(input$var_x, input$var_y)
    if (input$color_by != "none") vars <- c(vars, input$color_by)
    subset_data <- Diabetes_clean[, vars, drop = FALSE]

    cat("STATISTICAL SUMMARY\n")
    if (input$color_by != "none") {
      cat("Categorical variable:", input$color_by, "\n")
      print(table(subset_data[[input$color_by]]))
      cat("\n")
    }
    cat("Numeric variables:\n")
    print(summary(subset_data[, sapply(subset_data, is.numeric), drop = FALSE]))
  })
}

shinyApp(ui = ui, server = server)
```

---

## 7 — Conclusions

The analysis confirms that machine learning models can serve as useful
complementary tools for diabetes risk stratification. Both logistic regression
and Random Forest achieve strong discrimination (AUC > 0.87), with the latter
showing higher sensitivity (82.4%) for detecting positive cases.

**Key predictors**, ranked by Random Forest importance:
blood glucose (stab.glu, 24.3%), age (18.7%), BMI (15.2%), HDL (12.8%),
and waist-to-hip ratio (9.5%). This hierarchy validates the routine measurements
already available in pharmacy consultations and highlights the lipid profile as
a secondary but meaningful signal.

**Limitations** include the absence of lifestyle variables (diet, physical
activity, family history), a moderate sample size, and substantial missing data
in blood pressure variables. Future work should incorporate multiple imputation,
cross-validation, and external validation on independent cohorts to improve
generalisability.

| Model | Accuracy | Sensitivity | Specificity | AUC |
|---|---|---|---|---|
| Logistic Regression | 84.3% | 76.5% | 86.2% | 0.870 |
| Random Forest | 86.0% | 82.4% | 87.1% | 0.890 |
